Compositional Certified Resource Bounds\ifshort{\vspace{-1.15cm}}{}

### 2014-12-18 paper/main.tex
update logic rules and restore semantics paragraph

% \pref{fig:opsem} contains \ifshort{selected}{the} reduction rules of the semantics.  The
% rules define a rewrite system for program configurations of the form
% $(\state, S, \cont, \cost)$, where $\state$ is the program state, $S$
% is the statement being executed, $\cont$ is a continuation that
% describes what remains to be done after the execution of $S$, and
% $\cost \in \Q$ is the non-negative number of resources available for
% further execution.  All rules that can decrease $\cost$ have the
% implicit side condition that the resource quantity available
% \emph{before} the step is non-negative.  This means that we allow
% $(\state, S, \cont, \cost)$ with $c<0$ on the right-hand side
% transition relation $\smallstep{M}^*$ to indicate that the execution
% ran out of resources.  However, every execution that reaches such a
% state is stuck.
%
% \iffull{
% A continuation $\cont$ represents the context of the execution of a
% given statement.  A continuation can be the empty continuation
% $\Kstop$ (used to start a program), a sequenced computation $\Kseq S
% \cont$, a loop continuation $\Kloop S \cont$, or the continuation
% $\Kcall r \theta \cont$ of a function call.
% $$
% \cont := \Kstop
% \mid \Kseq S \cont
% \mid \Kloop S \cont
% \mid \Kcall r \theta \cont
% $$
% %
% During the execution we assume a fixed function context $\fenv$ that
% maps function names to a list of variables and the statement that
% defines the function body.  It is used in the rule \textsc{S:Call}.
% }{}
%
% The intuitive meaning of an evaluation \iffull{judgement}{} $(\state,
% S, \cont, \cost) \smallstep{M}^* (\state', S', \cont', \cost')$ is the
% following.  If the statement $S$ is executed in program state
% $\state$, with continuation $\cont$, and with $\cost$ resources
% available then---after a finite number of steps---the evaluation will
% reach the new machine state $(\state', S', \cont',\cost')$ and there
% are $\cost'$ resources available.  If $\cost' \ge 0$ then the
% execution did not run out of resources and the resource consumption up
% to this point is $\cost - \cost'$.  If this difference is negative
% then resource became available during the execution.  If however
% $\cost' < 0$ then the execution ran out of resources and is stuck.
% The cost of the execution is then $\cost \ge 0$.
%
% \iffull{
% \pref{lem:sem1} summarizes the main properties of the resource
% counter of the semantics.  It can be proved by induction on the number
% of steps.
% \begin{lemma}
% \label{lem:sem1}
% \itemskip
%   \begin{enumerate}
%   \item If $(\state, S, \cont, \cost) \smallstep{M}^k (\state', S',
%     \cont', \cost')$ and $n\ge 0$ then $(\state, S, \cont, \cost {+} n)
%     \smallstep{M}^k (\state', S', \cont', \cost' {+} n)$.
% \itemskipIn
%   \item If $(\state, S, \cont, \cost) \smallstep{M}^k (\state', S',
%     \cont', \cost')$ and $(\state, S, \cont, d) \smallstep{M}^k (\state', S',
%     \cont', d')$ then $\cost - \cost' = d - d'$.
%   \end{enumerate}
% \itemskip
% \end{lemma}
% %
% If we assume that we have a \emph{positive resource metric}, that is,
% a metric that assigns a positive cost to each evaluation step then we
% can link bounds on the resource consumption under the metric to
% termination.  This is formalized by \pref{lem:sem_term}.
% %
% \begin{lemma}
%  \label{lem:sem_term}
%  Let $M$ be a positive metric and assume that there exists $b \in
%  \Qplusz$ such that $(\state, S, \cont, \cost) \smallstep{M}^*
%  (\state', S', \cont', \cost') \implies \cost {-} \cost' {\leq} b$.
%  Then there exists $B \in \N$ such that $(\state, S, \cont, d)
%  \smallstep{M}^n (\state'', S', \cont', d')$ for some $n$, $d$, $d'$
%  and $\state''$ implies $n \le B$.
% \end{lemma}
% }

@@

In this article, we describe our system for a subset of Clight that is
sufficient to discuss the general ideas.  This subset is given by the
following grammar.
%
\begin{align*}
S &:= \code{assert}~E
\mid \code{skip}
\mid \code{break}
\mid \code{return}~x
\mid x \gets E
\mid x \gets f(x^*)
\\
& \mid \code{loop}~S
\mid \code{if}(E)~S~\code{else}~S
\mid S;S
\mid \code{tick}(n)
\end{align*}
%
Expressions $E$ are left abstract in our presentation.  For our
analysis framework, it is only important that they are side-effect
free.
%
The most notable difference to full Clight is that we can only assign
to variables and thus do not consider operations that update the heap.
Moreover, function arguments and return values are assumed to be
variables.  This is only for simplifying the presentation; in the
implementation we can deal with heap updates and general function
calls and returns.  However, we have not implemented our framework for function
pointers, \code{goto} statements, \code{continue} statements, and
\code{switch} statements.

We include the built-in primitive $\code{assert}~e$ that terminates
the program if the argument $e$ evaluates to false and has no effect
otherwise.  This is useful to express assumptions on the inputs of a
program for the automatic analysis.  We also add the built-in function
\code{tick(n)} that can be called with a constant integer $n$ as a
flexible way to model resource consumption or release (if $n$ is
negative).

\begin{figure*}[t!]
\begin{mathpar}
%
\Rule{S:Assert}
{ \code{istrue}~\sem{e}_\state }
{ (\state, \code{assert}~e, \cont, \cost)
\smallstep{M}
(\state, \code{skip}, \cont, \cost {-} M_a)
}
%
\and
\Rule{S:BrkSeq}
{}
{ (\state, \code{break}, \Kseq S \cont, \cost)
\smallstep{M}
(\state, \code{break}, \cont, \cost)
}
%
\and
\Rule{S:BrkLoop}
{}
{ (\state, \code{break}, \Kloop S \cont, \cost)
\smallstep{M}
(\state, \code{skip}, \cont, \cost {-} M_b)
}
%
\and
\Rule{S:RetSeq}
{}
{ (\state, \code{return}~x, \Kseq S \cont, \cost)
\smallstep{M}
(\state, \code{return}~x, \cont, \cost)
}
\and
%
\and
\Rule{S:RetLoop}
{}
{ (\state, \code{return}~x, \Kloop S \cont, \cost)
\smallstep{M}
(\state, \code{return}~x, \cont, \cost)
}
%
\and
\Rule{S:RetCall}
{ \state = (\_, \gamma) \\
  \state' = (\theta, \gamma)[r \mapsto \state(x)]
}
{ (\state, \code{return}~x, \Kcall r \theta \cont, \cost)
\smallstep{M}
(\state', \code{skip}, \cont, \cost {-} M_r)
}
%
\and
\Rule{S:Update}
{ \state' = \state[x \mapsto \sem{e}_\state] }
{ (\state, x \gets e, \cont, \cost)
\smallstep{M}
(\state', \code{skip}, \cont, \cost {-} M_u {-} M_e(e))
}
%
\and
\Rule{S:Call}
{ \fenv f = (\vec x, S_f) \\
  \state = (\theta, \gamma) \\
  \state' = (\vec x \mapsto \state(\vec y), \gamma) \\
}
{ (\state, r \gets f(\vec y), \cont, \cost)
\smallstep{M}
(\state', S_f, \Kcall r \theta \cont, \cost {-} M_f)
}
%
\and
\Rule{S:Loop}
{}
{ (\state, \code{loop}~S, \cont, \cost)
\smallstep{M}
(\state, S, \Kloop S \cont, \cost)
}
%
\and
\Rule{S:SkipLoop}
{}
{ (\state, \code{skip}, \Kloop S \cont, \cost)
\smallstep{M}
(\state, \code{loop}~S, \cont, \cost {-} M_l)
}
%
\and
\Rule{S:IfTrue}
{ \code{istrue}~\sem{e}_\sigma
\\ \cost' = \cost {-} M_c^1 {-} M_e(e)
}
{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
\smallstep{M}
(\state, S_1, \cont, \cost')
}
%
\and
\Rule{S:IfFalse}
{ \code{isfalse}~\sem{e}_\sigma
\\ \cost' = \cost {-} M_c^2 {-} M_e(e)
}
{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
\smallstep{M}
(\state, S_2, \cont, \cost')
}
%
\and
\Rule{S:Seq}
{ }
{ (\state, S_1; S_2, \cont, \cost)
\smallstep{M}
(\state, S_1, \Kseq {S_2} \cont, \cost)
}
%
\and
\Rule{S:SkipSeq}
{}
{ (\state, \code{skip}, \Kseq S \cont, \cost)
\smallstep{M}
(\state, S, \cont, \cost {-} M_s)
}
%
\and
\Rule{S:Tick}
{ }
{ (\state, \code{tick}(n), \cont, \cost)
\smallstep{M}
(\state, \code{skip}, \cont, \cost {-} M_t(n))
}

@@

@@

\begin{figure*}
\begin{mathpar}
%
\Rule{L:Skip}
{ }
{ \Delta; B; R \vdash
\htriple
  { Q }{ \code{skip} }{ Q }
}
%
\and
\Rule{L:Break}
{ }
{ \Delta; B; R \vdash
\htriple
  { M_b + B }{ \code{break} }{ Q }
}
%
\and
\Rule{L:Return}
{ }
{ \Delta; B; R \vdash
\htriple
  { R\,(\state(x)) }{ \code{return}~x }{ Q }
}
%
\and
\Rule[leftskip=.0cm,rightskip=.0cm]{L:Update}
{ }
{ \Delta; B; R \vdash
\htriple
  { \lambda \state.\,M_u + M_e(e) + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
}
%
\and
\Rule[leftskip=.0cm,rightskip=.0cm]{L:Seq}
{ \Delta; B; R \vdash
\htriple{ P }{ S_1 }{ Q' + M_s }
\\
\!\!  \Delta; B; R \vdash
\htriple{ Q' }{ S_2 }{ Q }
}
{ \Delta; B; R \vdash
\htriple{ P }{ S_1;S_2 }{ Q }
}
%
\and
\Rule{L:Call}
{
  \Delta(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v) \\
  P \models P_f \, y \, (\state(\vec x)) \land A \\
  \forall v.\, (Q_f \, y \, v \land A \models \lambda \state.\, Q \, \state[r \mapsto v])
}
{ \Delta; B; R \vdash
\htriple
  { M_f + P }
  { r \gets f(\vec x) }
  { Q - M_r }
}
%
\and
\Rule{L:Assert}
{ }
{ \Delta; B; R \vdash
\htriple
  { \code{istrue}~\sem{e}_\state \implies Q + M_a }
  { \code{assert}~e }
  { Q }
}
%
\and
\Rule{L:Tick}
{ }
{ \Delta; B; R \vdash
\htriple
  { Q + M_t(n) }
  { \code{tick}(n) }
  { Q }
}
%
\and
\Rule{L:Loop}
{ \Delta; Q; R \vdash
\htriple
  { I }{ S }{ I + M_l }
}
{ \Delta; B; R \vdash
\htriple
  { I }{ \code{loop}~S }{ Q }
}
%
\and
\Rule{L:If}
{ \Delta; B; R \vdash
\htriple
  { \code{istrue}~\sem{e}_\state + P - M_c^1}
  { S_1 }{ Q }
\\
  \Delta; B; R \vdash
\htriple
  { \code{isfalse}~\sem{e}_\state + P - M_c^2 }
  { S_2 }{ Q }
}
{ \Delta; B; R \vdash
\htriple
  { P + M_e(e) }{ \code{if}(e)~S_1~\code{else}~S_2 }{ Q }
}
%
\and
\Rule{L:Weaken}
{
  \!\! P \models P' \!\! \\
  \!\! \Delta; B'; R' \vdash
  \htriple{ P' }{ S }{ Q' } \!\! \\
  \!\! Q' \models Q \!\! \\
  \!\! B' \models B \!\! \\
  \!\! \forall v.\, (R'\,v \models R\,v) \!\!
}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q }
}
%
\and
\Rule{L:Frame}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q } \\
  x \in \Qplusz
}
{ \!\! \Delta; B + x; R + x \vdash
\htriple{ P + x }{ S }{ Q + x } \!\!
}
%
\and
\Rule{L:Extend}
{ \Delta \cup \Delta'; B; R \vdash
  \htriple{ P }{ S }{ Q }
\\
  \forall f\,P_f\,Q_f.\,
  \Delta'(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v)
  \rightarrow
  \forall y\,\vec v.\,
  (\Delta \cup \Delta'; \bot; Q_f\,y \vdash
    \htriple{ P_f\,y\,\vec v }{ S_f }{ \bot })
}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q }
}
%
\end{mathpar}

@@

In this section we describe a simplified version of the quantitative
Hoare logic that we use in Coq to interactively prove resource bounds.
%
We generalize classic Hoare logic to express not only classical
boolean-valued assertions but also assertions that talk about the
future resource usage.  Instead of the usual assertions $P : \State
\to \mathit{bool}$ of Hoare logic we use assertions
$$
P : \State \to \Qplusz \cup \{ \infty \} \; .
$$
This can be understood as a refinement of boolean assertions where
$\mathit{false}$ is $\infty$ and $\mathit{true}$ is refined by $\Qplusz$.
We write $\Assn$ for $\State \to \Qplusz \cup \{ \infty \}$ and $\bot$ for
$\lambda \state . \, \infty$.  We sometimes call assertions
\emph{potential functions}.  To use Coq's support for propositional
reasoning, assertions have the type $\State \to \Qplusz \to \mathrm{Prop}$
in the implementation.  For a given $\state \in \State$, such an
assertion can be seen as a set $B \subseteq \Qplusz$ of valid bounds.
However, we find the presentation in this article easier to read.

Due to \code{break} and \code{return} statements of Clight, there are
different possible ways to exit a block of code.  We also have to keep
track of the resource specifications of functions.  To account for
this in the logic, our quantitative Hoare triples have the form
$$
\Delta; B; R \vdash
\htriple
  { Q }
  { S }
  { Q' } \; .
$$
The triple $\htriple{ Q }{ S }{ Q' }$ consists of a statement $S$ and
two assertions $Q,Q' : \Assn$.  It corresponds to triples in classic
Hoare logic and the intuitive meaning is as follows.  If $S$ is
executed with starting state $\state$, the empty continuation
$\Kstop$, and at least $P(\state)$ resources available then the
evaluation does not run out of resources and there are at least
$Q(\state')$ resources left if the evaluation terminates in state
$\state'$.  The assertion $B : \Assn$ provides the
postcondition for the case in which the code block $S$ is exited by a
\code{break} statement.  So if the execution is terminated in state
$\state'$ with a \code{break} then $B(\state')$ resources are
available.  Similarly, $R : \Z \to \Assn$ is the postcondition for the
case in which the code block $S$ is exited by a $\code{return}~x$
statement.  The integer argument of $R$ is the return value.
Finally, the function context of judgements that we write $\Delta$ is
a mapping from function names to specifications of the form
$$
  \forall z \, \vec v \, v .(P_f \, z \, \vec v, Q_f \, z \, v).
$$
The assertion $P_f \, z \, \vec v$ is the precondition of the function
$f$ and the assertion $Q_f \, z \, v$ is its postcondition.  They are
both parameterized by an arbitrary logical variable $z$ (which can be a
tuple) that relates the function arguments with the return value.  The
precondition also depends on $\vec v$, the values of the arguments
at the function invocation.  Similarly, the postcondition depends on
the return value $v$ of the function.  The use of logical variables to
express relations between different states of an execution is a
standard technique of Hoare logic.
%
To ensure soundness, we require that $P_f$ and $Q_f$ do not depend on
the local variables on the stack, that is, $\forall z \, \vec v
\, \theta \, \theta' \, \gamma . \, P_f \, z \, \vec v \,
(\theta,\gamma) = P_f \, z \, \vec v \, (\theta',\gamma)$.

For two assertions $P,Q : \Assn$, we write $P \models Q$ to if for all
program state $\state$ $P(\state) \ge Q(\state)$.

@@

Finally, we describe the rules which are not syntax directed.  There
are two weakening rules available in the quantitative Hoare logic.
The framing rule {\sc L:Frame} is designed to weaken a statement by
stating that if $S$ needs $P$ resources to run and leaves $Q$
resources available after its execution, then it can very well run
with $P + c$ resources and return $Q + c$ resources.  The consequence
{\sc L:Conseq} rule is directly imported from classical Hoare logic
except that instead of using the logical implication $\Rightarrow$ we
use the quantitative $\models$ that point-wise applies $\ge$.  This
rule indeed weakens the statement since it requires more resource to
run the statement and yields less than what has been proved to be
available after its termination.

@@

We already gave an intuition of the meaning of judgements
derived in the logic.  To make it formal, we define the
\emph{resource safety} $\safe n P S \cont$ of an assertion
and a program configuration as $\forall \state \, \cost \, m \, \cost' .$
\begin{align*}
  (m {\le} n \land P(\state) {\le} \cost \land
    (\state, S, \cont, \cost) \smallstep{M}^m (\_, \_, \_, \cost'))
  \implies \cost' {\ge} 0.
\end{align*}
This predicate is step indexed by an integer $n$ that is used for
induction in the soundness proof for the function-call and loop cases.
The constraint $\cost' \geq 0$ imposed by the definition ensures that
the program does not stop because of a resource error (recall that a
negative resource counter on the right-hand side is a resource
failure).  However, it does not rule out memory safety errors or
assertion failures. This is because our logic does not prove any
safety or correctness theorems but only focuses on resource usage.
%
An interesting detail in the definition is the natural number $m$.  We
simply use it to ensure that $\safe {n+1} P S \cont \implies \safe n P S
\cont $.  This would not be the case if we replaced all occurrences of
$m$ by $n$ in the definition of validity.

>>>>

@@

We describe our system for a subset of Clight that is sufficient to
discuss the general ideas.  This subset is given by the following
grammar.
%
\begin{align*}
S &:= \code{assert}~E
\mid \code{skip}
\mid \code{break}
\mid \code{return}~x
\mid x \gets E
\mid x \gets f(x^*)
\\
& \mid \code{loop}~S
\mid \code{if}(E)~S~\code{else}~S
\mid S;S
\mid \code{tick}(n)
\end{align*}
%
Expressions $E$ are left abstract in our presentation.  For our
analysis framework, it is only important that they are side-effect
free.
%
The most notable difference to full Clight is that we can only assign
to variables and thus do not consider operations that update the heap
or the stack through a pointer.
Moreover, function arguments and return values are assumed to be
variables.  This is only for simplifying the presentation; in the
implementation we can deal with heap updates and general function
calls and returns.  However, we have not implemented our framework for function
pointers, \code{goto} statements, \code{continue} statements, and
stack variables that have their address taken.

We include the built-in primitive $\code{assert}~e$ that terminates
the program if the argument $e$ evaluates to false and has no effect
otherwise.  This is useful to express assumptions on the inputs of a
program for the automatic analysis.  We also add the built-in function
\code{tick(n)} that can be called with a constant rational number $n$
as a flexible way to model resource consumption or release (if $n$ is
negative).

\begin{figure*}[t!]
\begin{mathpar}
%
\Rule{S:Assert}
{ \code{istrue}~\sem{e}_\state }
{ (\state, \code{assert}~e, \cont, \cost)
\smallstep
(\state, \code{skip}, \cont, \cost)
}
%
\and
\Rule{S:BrkSeq}
{ }
{ (\state, \code{break}, \Kseq S \cont, \cost)
\smallstep
(\state, \code{break}, \cont, \cost)
}
%
\and
\Rule{S:BrkLoop}
{ }
{ (\state, \code{break}, \Kloop S \cont, \cost)
\smallstep
(\state, \code{skip}, \cont, \cost)
}
%
\and
\Rule{S:RetSeq}
{ }
{ (\state, \code{return}~x, \Kseq S \cont, \cost)
\smallstep
(\state, \code{return}~x, \cont, \cost)
}
\and
%
\and
\Rule{S:RetLoop}
{ }
{ (\state, \code{return}~x, \Kloop S \cont, \cost)
\smallstep
(\state, \code{return}~x, \cont, \cost)
}
%
\and
\Rule{S:RetCall}
{ \state = (\_, \gamma) \\
  \state' = (\theta, \gamma)[r \mapsto \state(x)]
}
{ (\state, \code{return}~x, \Kcall r \theta \cont, \cost)
\smallstep
(\state', \code{skip}, \cont, \cost)
}
%
\and
\Rule{S:Update}
{ \state' = \state[x \mapsto \sem{e}_\state] }
{ (\state, x \gets e, \cont, \cost)
\smallstep
(\state', \code{skip}, \cont, \cost))
}
%
\and
\Rule{S:Call}
{ \fenv f = (\vec x, S_f) \\
  \state = (\theta, \gamma) \\
  \state' = (\vec x \mapsto \state(\vec y), \gamma) \\
}
{ (\state, r \gets f(\vec y), \cont, \cost)
\smallstep
(\state', S_f, \Kcall r \theta \cont, \cost)
}
%
\and
\Rule{S:Loop}
{ }
{ (\state, \code{loop}~S, \cont, \cost)
\smallstep
(\state, S, \Kloop S \cont, \cost)
}
%
\and
\Rule{S:SkipLoop}
{ }
{ (\state, \code{skip}, \Kloop S \cont, \cost)
\smallstep
(\state, \code{loop}~S, \cont, \cost)
}
%
\and
\Rule{S:IfTrue}
{ \code{istrue}~\sem{e}_\sigma
}
{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
\smallstep
(\state, S_1, \cont, \cost)
}
%
\and
\Rule{S:IfFalse}
{ \code{isfalse}~\sem{e}_\sigma
}
{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
\smallstep
(\state, S_2, \cont, \cost)
}
%
\and
\Rule{S:Seq}
{ }
{ (\state, S_1; S_2, \cont, \cost)
\smallstep
(\state, S_1, \Kseq {S_2} \cont, \cost)
}
%
\and
\Rule{S:SkipSeq}
{ }
{ (\state, \code{skip}, \Kseq S \cont, \cost)
\smallstep
(\state, S, \cont, \cost)
}
%
\and
\Rule{S:Tick}
{ }
{ (\state, \code{tick}(n), \cont, \cost)
\smallstep
(\state, \code{skip}, \cont, \cost - n)
}

@@

\paragraph{Semantics.}

\pref{fig:opsem} contains the reduction rules of the semantics.  The rules
define a rewrite system for program configurations of the form $(\state,
S, \cont, \cost)$, where $\state$ is the program state, $S$ is the
statement being executed, $\cont$ is a continuation that describes what
remains to be done after the execution of $S$, and $\cost \in \Q$ is
the amount of resources available for further execution.  All the rules
have the implicit side condition that the resource quantity available
\emph{before} the step is non-negative.  This means that we allow
$(\state, S, \cont, \cost)$ with $c<0$ on the right-hand side transition
relation $\smallstep^*$ to indicate that the execution ran out of
resources.  However, every execution that reaches such a state is stuck.

A continuation $\cont$ represents the context of the execution of a
given statement.  A continuation can be the empty continuation
$\Kstop$ (used to start a program), a sequenced computation $\Kseq S
\cont$, a loop continuation $\Kloop S \cont$, or the continuation
$\Kcall r \theta \cont$ of a function call.
$$
\cont := \Kstop
\mid \Kseq S \cont
\mid \Kloop S \cont
\mid \Kcall r \theta \cont
$$
%
During the execution we assume a fixed function context $\fenv$ that
maps function names to a list of variables and the statement that
defines the function body.  It is used in the rule \textsc{S:Call}.

The intuitive meaning of an evaluation $(\state,
S, \cont, \cost) \smallstep^* (\state', S', \cont', \cost')$ is the
following.  If the statement $S$ is executed in program state
$\state$, with continuation $\cont$, and with $\cost$ resources
available then---after a finite number of steps---the evaluation will
reach the new machine state $(\state', S', \cont',\cost')$ and there
are $\cost'$ resources available.  If $\cost' \ge 0$ then the
execution did not run out of resources and the resource consumption up
to this point is $\cost - \cost'$.  If this difference is negative
then resource became available during the execution.  If however
$\cost' < 0$ then the execution ran out of resources and is stuck.
The cost of the execution is then $\cost \ge 0$.

\pref{lem:sem1} summarizes the main properties of the resource
counter of the semantics.  It can be proved by induction on the number
of steps.
\begin{lemma}
\label{lem:sem1}
\itemskip
  \begin{enumerate}
  \item If $(\state, S, \cont, \cost) \smallstep^k (\state', S',
    \cont', \cost')$ and $n\ge 0$ then $(\state, S, \cont, \cost {+} n)
    \smallstep^k (\state', S', \cont', \cost' {+} n)$.
\itemskipIn
  \item If $(\state, S, \cont, \cost) \smallstep^k (\state', S',
    \cont', \cost')$ and $(\state, S, \cont, d) \smallstep^k (\state', S',
    \cont', d')$ then $\cost - \cost' = d - d'$.
  \end{enumerate}
\itemskip
\end{lemma}

% OOPS this is no longer true... This is an argument agains a presentation
% with ticks only. XXX
%
% If we assume that we have a \emph{positive resource metric}, that is,
% a metric that assigns a positive cost to each evaluation step then we
% can link bounds on the resource consumption under the metric to
% termination.  This is formalized by \pref{lem:sem_term}.
%
% \begin{lemma}
%  \label{lem:sem_term}
%  Let $M$ be a positive metric and assume that there exists $b \in
%  \Qplusz$ such that $(\state, S, \cont, \cost) \smallstep^*
%  (\state', S', \cont', \cost') \implies \cost {-} \cost' {\leq} b$.
%  Then there exists $B \in \N$ such that $(\state, S, \cont, d)
%  \smallstep^n (\state'', S', \cont', d')$ for some $n$, $d$, $d'$
%  and $\state''$ implies $n \le B$.
% \end{lemma}

@@

\begin{figure*}
\begin{mathpar}
%
\Rule{L:Skip}
{ }
{ \Delta; B; R \vdash
\htriple
  { Q }{ \code{skip} }{ Q }
}
%
\and
\Rule{L:Break}
{ }
{ \Delta; B; R \vdash
\htriple
  { B }{ \code{break} }{ Q }
}
\and
\Rule{L:Tick}
{ n < 0 \implies Q \ge 0}
{ \Delta; B; R \vdash
\htriple
  { Q }
  { \code{tick}(n) }
  { Q - n }
}
%
%
\and
\Rule{L:Return}
{ }
{ \Delta; B; R \vdash
\htriple
  { R\,(\state(x)) }{ \code{return}~x }{ Q }
}
%
\and
\Rule{L:Assert}
{ }
{ \Delta; B; R \vdash
\htriple
  { \code{istrue}~\sem{e}_\state \implies Q }
  { \code{assert}~e }
  { Q }
}
%
\and
\Rule{L:Update}
{ }
{ \Delta; B; R \vdash
\htriple
  { \lambda \state.\, Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
}
%
\and
\Rule{L:Loop}
{ I \ge Q
\\ \Delta; Q; R \vdash
   \htriple { I }{ S }{ I }
}
{ \Delta; B; R \vdash
\htriple
  { I }{ \code{loop}~S }{ Q }
}
%
\and
\Rule{L:If}
{ \Delta; B; R \vdash
\htriple
  { \code{istrue}~\sem{e}_\state + P }
  { S_1 }{ Q }
\\\\
  \Delta; B; R \vdash
\htriple
  { \code{isfalse}~\sem{e}_\state + P }
  { S_2 }{ Q }
}
{ \Delta; B; R \vdash
\htriple
  { P }{ \code{if}(e)~S_1~\code{else}~S_2 }{ Q }
}
%
\and
\Rule{L:Seq}
{ \Delta; B; R \vdash
\htriple{ P }{ S_1 }{ Q' }
\\
  \Delta; B; R \vdash
\htriple{ Q' }{ S_2 }{ Q }
}
{ \Delta; B; R \vdash
\htriple{ P }{ S_1;S_2 }{ Q }
}
%
\and
\Rule{L:Call}
{
  \Delta(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v) \\
  P \ge P_f \, y \, (\state(\vec x)) + A \\
  \forall v.\, (Q_f \, y \, v + A \ge \lambda \state.\, Q \, \state[r \mapsto v])
}
{ \Delta; B; R \vdash
\htriple
  { P }
  { r \gets f(\vec x) }
  { Q }
}
%
\and
\Rule{L:Extend}
{  \Delta \cup \Delta'; B; R \vdash
  \htriple{ P }{ S }{ Q }
\\
  P_f \ge 0
\\\\
  \forall f\,P_f\,Q_f.\,
  \Delta'(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v)
  \rightarrow
  \forall y\,\vec v.\,
  (\Delta \cup \Delta'; \bot; Q_f\,y \vdash
    \htriple{ P_f\,y\,\vec v }{ S_f }{ \bot })
}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q }
}
%
\and
\Rule{L:Weaken}
{
  \Delta; B'; R' \vdash \htriple{ P' }{ S }{ Q' } \\\\
  P \ge P' \!\! \\
  \!\! Q' \ge Q \!\! \\
  \!\! B' \ge B \!\! \\
  \!\! \forall v.\, (R'\,v \ge R\,v)
}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q }
}
%
\and
\Rule{L:Frame}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q } \\
  x \in \Q
}
{ \!\! \Delta; B + x; R + x \vdash
\htriple{ P + x }{ S }{ Q + x } \!\!
}
%
\end{mathpar}

@@

In this section we describe a simplified version of the quantitative
Hoare logic that we use in Coq to interactively prove resource bounds.
%
We generalize classic Hoare logic to express not only classical
boolean-valued assertions but also assertions that talk about the
future resource usage.  Instead of the usual assertions $P : \State
\to \mathit{bool}$ of Hoare logic we use assertions
$$
P : \State \to \Q \cup \{ \infty \} \; .
$$
This can be understood as a refinement of boolean assertions where
$\mathit{false}$ is $\infty$ and $\mathit{true}$ is refined by $\Qplusz$.
We write $\Assn$ for $\State \to \Q \cup \{ \infty \}$ and $\bot$ for
$\lambda \state . \, \infty$.  We sometimes call assertions
\emph{potential functions}.  To use Coq's support for propositional
reasoning, assertions have the type $\State \to \Q \to \mathrm{Prop}$
in the implementation.  For a given $\state \in \State$, such an
assertion can be seen as a set $B \subseteq \Q$.  However, we find
the presentation in this article easier to read.

Due to \code{break} and \code{return} statements of Clight, there are
different possible ways to exit a block of code.  We also have to keep
track of the resource specifications of functions.  To account for
this in the logic, our quantitative Hoare triples have the form
$$
\Delta; B; R \vdash
\htriple
  { Q }
  { S }
  { Q' } \; .
$$
The triple $\htriple{ Q }{ S }{ Q' }$ consists of a statement $S$ and
two assertions $Q,Q' : \Assn$.  It corresponds to triples in classic
Hoare logic and the intuitive meaning is as follows.  If $S$ is
executed with starting state $\state$, with the empty continuation
$\Kstop$, $Q' \ge 0$, and at least $Q(\state)$ resources available then the
evaluation does not run out of resources and there are at least
$Q'(\state')$ resources left if the evaluation terminates in state
$\state'$.  The assertion $B : \Assn$ provides the
postcondition for the case in which the code block $S$ is exited by a
\code{break} statement.  So if the execution is terminated in state
$\state'$ with a \code{break} then $B(\state')$ resources are
available.  Similarly, $R : \Z \to \Assn$ is the postcondition for the
case in which the code block $S$ is exited by a $\code{return}~x$
statement.  The integer argument of $R$ is the return value.
Finally, the function context of judgements that we write $\Delta$ is
a mapping from function names to specifications of the form
$$
  \forall z \, \vec v \, v .(P_f \, z \, \vec v, Q_f \, z \, v).
$$
The assertion $P_f \, z \, \vec v$ is the precondition of the function
$f$ and the assertion $Q_f \, z \, v$ is its postcondition.  They are
both parameterized by an arbitrary logical variable $z$ (which can be a
tuple) that relates the function arguments with the return value.  The
precondition also depends on $\vec v$, the values of the arguments
at the function invocation.  Similarly, the postcondition depends on
the return value $v$ of the function.  The use of logical variables to
express relations between different states of an execution is a
standard technique of Hoare logic.
%
To ensure soundness, we require that $P_f$ and $Q_f$ do not depend on
the local variables on the stack, that is, $\forall z \, \vec v
\, \theta \, \theta' \, \gamma . \, P_f \, z \, \vec v \,
(\theta,\gamma) = P_f \, z \, \vec v \, (\theta',\gamma)$.

For two assertions $P,Q : \Assn$, we write $P \ge Q$ to if for all
program state $\state$, $P(\state) \ge Q(\state)$.

@@

Finally, we describe the rules which are not syntax directed.  There
are two weakening rules available in the quantitative Hoare logic.
The framing rule {\sc L:Frame} is designed to weaken a statement by
stating that if $S$ needs $P$ resources to run and leaves $Q$
resources available after its execution, then it can very well run
with $P + c$ resources and return $Q + c$ resources.  The consequence
{\sc L:Conseq} rule is directly imported from classical Hoare logic
except that instead of using the logical implication $\Rightarrow$ we
use the quantitative $\ge$.  This rule indeed weakens the statement
since it requires more resource to run the statement and yields less
than what has been proved to be available after its termination.

@@

We already gave an intuition of the meaning of judgements
derived in the logic.  To make it formal, we define the
\emph{resource safety} $\safe n P S \cont$ of an assertion
and a program configuration as $\forall \state \, \cost \, m \, \cost' .$
\begin{align*}
  (m {\le} n \land P(\state) {\le} \cost \land
    (\state, S, \cont, \cost) \smallstep^m (\_, \_, \_, \cost'))
  \implies \cost' {\ge} 0.
\end{align*}
This predicate is step indexed by an integer $n$ that is used for
induction in the soundness proof for the function-call and loop cases.
The constraint $\cost' \geq 0$ imposed by the definition ensures that
the program does not stop because of a resource error (recall that a
negative resource counter on the right-hand side is a resource
failure).  However, it does not rule out memory safety errors or
assertion failures. This is because our logic does not prove any
safety or correctness theorems but only focuses on resource usage.
%
An interesting detail in the definition is the natural number $m$.  We
simply use it to ensure that $\safe {n+1} P S \cont \implies \safe n P S
\cont $.  This would not be the case if we replaced all occurrences of
$m$ by $n$ in the definition of validity.
